Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-01 ind_files/03_field_eqns.tex
Remove unused files, renumber filenames so they still make sense

\section{The infinite population limit}
We can once again appeal to the link between Fokker-Planck equations and Langevins \citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where we have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE' models discussed in chapter \ref{chap_AD}. They are also equivalent to the `oligomorphic dynamics' of \citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology \citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of \citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics.\\
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear. Nevertheless, we will pretend all is well and assume that one can do this, bolstered by the fact that we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, as we show below.

@@

Note that by definition, $B(y(t))$ is the variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} represents a broad class of systems called `gradient equations' in quantitative genetics \citep{lande_natural_1976, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  Comparing with \eqref{AD_canonical_eqn}, we can see that this is very similar to the canonical equation of adaptive dynamics, and indeed, the two are deeply related \citep{lehtonen_price_2018, lion_theoretical_2018}. The major difference is that in the `proper' canonical equation of adaptive dynamics (as formulated in \cite{dieckmann_dynamical_1996}), the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory \citep{champagnat_unifying_2006}. A heuristic derivation of the canonical equation of adaptive dynamics is provided in the classic article by \cite{dieckmann_dynamical_1996}.

>>>>

\section{The infinite population limit}
We can once again appeal to the link between Fokker-Planck equations and Langevins \citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where we have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE' models discussed in studies of Adaptive Diversification \citep{doebeli_adaptive_2011}. They are also equivalent to the `oligomorphic dynamics' of \citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology \citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of \citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics.\\
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear. Nevertheless, we will pretend all is well and assume that one can do this, bolstered by the fact that we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, as we show below.

@@

Note that by definition, $B(y(t))$ is the variance of the trait in the population at time $t$. The term $w\left(z|N_K\delta_{y(t)}\right)$ is the expected growth rate of an individual with trait value $z$ in a population of size $N_K$ in which (almost) every individual has trait value $y$. This quantity is referred to as the `invasion fitness' of a `mutant' trait $z$ in a population of `resident' $y$ individuals. Equation \eqref{AD_canonical_derived} represents a broad class of systems called `gradient equations' in quantitative genetics \citep{lande_natural_1976, abrams_relationship_1993, lehtonen_price_2018, lion_theoretical_2018}, and captures the approximate evolutionary dynamics of quantitative traits under certain mutation limits.  It is also deeply related \citep{lehtonen_price_2018, lion_theoretical_2018} to the canonical equation of adaptive dynamics \citep{dieckmann_dynamical_1996, doebeli_adaptive_2011}. The major difference is that in the `proper' canonical equation of adaptive dynamics (as formulated in \cite{dieckmann_dynamical_1996}), the function $B(y)$ explicitly relies on mutations as a continual source of variation, whereas in gradient dynamics and our equation \eqref{AD_canonical_derived}, $B(y)$ captures the standing genetic variation in the population but does not specify the source of this variation. Note that strictly speaking, if $\psi(x,t) = \delta_{y(t)}$ exactly, then $B(y) \equiv 0$. This just reflects our assumption that mutations are vanishingly rare and mutants are sampled from infinitesimally close to the resident value. More detailed mathematical arguments are required to ensure that this convergence `makes sense' and that $B(y)$ does not actually equal 0. This has been proved rigorously using much more sophisticated mathematical tools grounded in martingale theory \citep{champagnat_unifying_2006}. A heuristic derivation of the canonical equation of adaptive dynamics is provided in the classic article by \cite{dieckmann_dynamical_1996}.
